---
title: "Data Visualization"
---

```{r, message=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(collapse = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggfortify)
library(pheatmap)
theme_set(theme_bw())
```


## Overview  

- Effective visualizations
- Grammar of graphics
- ggplot practical

## Effective visualizations

- Why do we plot data?
- What makes a good plot?

## Plot types

-----  

### Scatter plot

```{r, echo=FALSE}
ggplot(mtcars, aes(wt, mpg)) + geom_point() +
	labs(x = "Weight (1000 lbs)", y = "Miles/(US) gallon", caption = "Figure 1. Fuel efficiency decreases with increased vehicle weight")
```


-----  

### Bar chart

```{r, echo=FALSE}
ggplot(mpg, aes(class)) + geom_bar() + labs(x = "Vehicle Class")
```


-----  

### Histogram

```{r, echo=FALSE, message=FALSE}
ggplot(diamonds, aes(price)) + geom_histogram() + labs(x = "Diamond Price ($US)", title = "Summary of diamond prices")
```


------  

### Boxplot

```{r, echo=FALSE}
ggplot(chickwts, aes(feed, weight)) + geom_boxplot() + 
	labs(x = "Feed Type", y = "Weight (g)", title = "Effect of feed type on chick weight")
```

-----  

### Line plots
```{r, echo=FALSE}

DNase %>% group_by(conc) %>% summarise(mean = mean(density), sd = sd(density)) %>% 
	ggplot(aes(conc, mean)) + 
	geom_errorbar(aes(ymax = mean + sd, ymin = mean - sd)) + 
	geom_point() + geom_line() +
	labs(x = "DNase Concentration (mg/mL)", y = "Mean optical density")

```

-----  

### Time series
```{r, echo=FALSE}
autoplot(lynx) + labs(x = "Year", y = "Number of lynx trappings", title = "Annual Canadian Lynx trappings")
```


-----  

### Heatmap

```{r, echo=FALSE}
mat = matrix(rnorm(1:100), nrow = 20)
pheatmap(mat)
```



## Good Visualizations 

- tell a story
- are easy to understand
- are simple
- contain all the information need to interpret the plot


## Tips

- Think about what you are trying to convey
- Don't include a figure because it looks "cool"
- Ask yourself if the figure is as simple as it could be
- Use colours but think about color blindness
- Don't use more than 10 - 12 colours to distinguish groups
- Sometimes a table is better

----- 

- Don't use pie charts
- Don't use stacked bar charts
- Use facet panels to reduce clutter
- Avoid overplotting

-----

See the book, _The Visual Display of Quantitative Information_ by Edward Tufte for more information.


## Grammar of Graphics

- A vocabulary to descibe how to build a statistical graphic.  

- Plot elements are well defined and by combining them using a set of rules we can create a wide variety of complex statistical graphics.

- We build graphics like we build sentences and paragraphs.


## More info

ggplot2 [book](https://www.amazon.ca/ggplot2-Elegant-Graphics-Data-Analysis/dp/331924275X/ref=sr_1_1?ie=UTF8&qid=1474233790&sr=8-1&keywords=elegant+graphics+for+data+analysis)

[Cookbook for R](http://www.cookbook-r.com/Graphs/)  

ggplot2 [documentation](http://docs.ggplot2.org/current/)  



