---
title: "Data Visualization"
---

# Plots, looping over plots, and creating a plotting function
For loops

First, let's organize our data and get the code to work for one of the items we want to loop through. 


```{r, echo=FALSE, results='hide'}


```


## Plot for a single compound
```{r, echo=FALSE}


```

Yey! Our single-compound plot works! Now, let's make it a for loop

## Convert the single-compound plot into a for loop (automating)  

Frequently enough, we find cases where we need to do several similar graphs from one or several datasets, following the same structure. In those cases, we have two options: copy, paste and ajust variables n times, or script a flow control method (for loops, apply family, etc.) to achieve our goal. The code bellow loops over our data to creates plots.


```{r, echo=FALSE}


```

Now, your labmates/coworkers/collegues want the same type of plots you created, but they don't know how to do it. What would you do after making the plots for one, then two peers?

Create a function and pass it along (meaning, "good luck!")

```{r, echo=FALSE}


```

[for loops for plots](http://www.reed.edu/data-at-reed/resources/R/loops_with_ggplot2.html)


```{r, echo=FALSE}
sessionInfo()

```

