---
title: "Basic Data Analysis"
---

```{r, message=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(collapse = TRUE, message = FALSE, results = 'hide')
options("readr.num_columns" = 0)
```


```{r}
library("dplyr")
library("tidyr")
library("readr")
```

## Load our demo dataset
```{r}
pa = read_csv("../datasets/pa_phenotype_small.csv")
```

What does it look like?
```{r}
glimpse(pa)
```

## Tidy

Let's tidy it up a bit. The `gather` function is what we use to gather up multiple columns (observations) and put them into one column.

Here we need to decide which of our columns contain the values we're measuring.
```{r}
pa_tidy = pa %>% 
	gather(Phenotype, Measurement, -Strain_ID, -Event_num, -Clinical_Status, -Date_of_Isolation)
```

Spread is the reverse of gather
```{r}
pa_tidy %>% spread(Phenotype, Measurement)
```


## Dplyr verbs

### Summarise

The `summarise` function is used to create statistical summaries of your data.  Most of the time it is used in combination with `group_by`.  Let's see how this works.

Let's find the mean of each phenotype 
```{r}
pa_tidy %>% 
	group_by(Phenotype) %>% 
	summarise(mean = mean(Measurement, na.rm = TRUE))

```

Hmm, how does that look if we divide it up by clinical status?

```{r}
pa_tidy %>% 
	group_by(Phenotype, Clinical_Status) %>% 
	summarise(mean = mean(Measurement, na.rm = TRUE))

```

What about other stats?
```{r}
pa_tidy %>% 
	group_by(Phenotype) %>% 
	summarise(mean = mean(Measurement, na.rm = TRUE),
						sd = sd(Measurement, na.rm = TRUE),
						max = max(Measurement, na.rm = TRUE),
						min = min(Measurement, na.rm = TRUE))

```

### Mutate

Mutate is similar to summarise, however, it "mutates" a column instead of summarising it.  Summarise takes a bunch of values and produces on value from those.  Mutate returns a transformed vector that is the same size as the original. Usually we're not grouping but you can if needed.

Let's use mutate here to transform one of our variables to log scale
```{r}
pa_log = pa %>% mutate(Pyochelin_trans = log10(Pyochelin))
```

Look at the transformed variable.  What's the problem here?
```{r}
pa_log$Pyochelin_trans
```

```{r}
pa_log = pa %>% mutate(Pyochelin_trans = log10(Pyochelin + 1))
```

Alternative way to do the same thing, although note that this modifies our dataframe in-place instead of returning a copy.

```{r}
pa$Pyochelin_trans = log10(pa$Pyochelin + 1)
```

Transmute does the same as mutate but returns only the mutated column
```{r}
pa %>% transmute(Pyochelin_trans = log10(Pyochelin + 1))
```


### Filter

`filter` returns only rows matching a particular condition
```{r}
pa_tidy %>% filter(Phenotype == "CAZ")
pa_tidy %>% filter(Date_of_Isolation < "2008-01-01")
pa_tidy %>% filter(Phenotype == "CAZ", Measurement >= 20)
```

In base R:
```{r}
pa_tidy[pa_tidy$Phenotype == "CAZ" & pa_tidy$Measurement >= 20, ]
```


### Selecting columns

`select` gives us the columns we want. Use unqouted column names.

```{r}
pa %>% select(Strain_ID, CIP, Swim)
```

Columns are returned in the order we list them
```{r}
pa %>% select(Date_of_Isolation, LB, Strain_ID, Swim, Swarm)
```

Dplyr provides helper functions for selecting columns. `?select_helpers`
```{r}
pa %>% select(everything())
pa %>% select(starts_with("P"))
pa %>% select(ends_with("Z"))
pa %>% select(contains("A"))

# Move a column to the front
pa %>% select(Swarm, everything())
```

### Other "verbs"

`slice` gives us row indexing
```{r}
pa %>% 
	select(Strain_ID, CIP, Swim) %>% 
	slice(1:10)
```

`arrange` orders the dataframe by a variable
```{r}
pa %>% arrange(Date_of_Isolation)
pa %>% arrange(desc(Date_of_Isolation))
```

`distinct` gives us only the unique rows
```{r}
pa_tidy %>% distinct(Event_num)
pa_tidy %>% distinct(Date_of_Isolation)
```

`count` and `tally` count occurances
```{r}
pa_tidy %>% count(Date_of_Isolation)
pa_tidy %>% group_by(Phenotype, Date_of_Isolation) %>% tally()
```

`sample_n` and `sample_frac` allow us to randomly sample rows
```{r}
pa %>% sample_n(5)
pa %>% sample_frac(0.2)
```


## Putting it together

What are the mean antibiotic scores for each event?
```{r}
pa_tidy %>% 
	filter(Phenotype %in% c("CAZ", "CIP", "CT")) %>% 
	group_by(Event_num, Phenotype) %>% 
	summarise(mean_antibiotic = mean(Measurement, na.rm = TRUE)) %>% 
	arrange(mean_antibiotic)
```


Mean center (scale) the phenotype measurments.
```{r}
pa_tidy %>% 
	group_by(Phenotype) %>% 
	mutate(centered = scale(Measurement))
```

Tidy, log transorm the Pyochelin column, and show the difference of the medians for each phenotype between clinically stable and exacerbation.  Then add a column to indicate if the change is positve or negative.

```{r}
pa %>% 
	gather(Phenotype, Measurement, -Strain_ID, -Event_num, -Clinical_Status, -Date_of_Isolation) %>% 
	mutate(Measurement = if_else(Phenotype == "Pyochelin", log10(Measurement), Measurement)) %>% 
	group_by(Phenotype, Clinical_Status) %>% 
	summarise(median = median(Measurement, na.rm = TRUE)) %>% 
	spread(Clinical_Status, median) %>% 
	mutate(difference = Exacerbation - `Clinically Stable`) %>% 
	mutate(Pos_or_neg = if_else(sign(difference) == -1, "-", "+"))
```






