---
title: "Tidy gene expression and heatmap"
output: pdf_document
---

```{r, message=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(collapse = TRUE) 
```

```{r}
library(tidyr)
library(dplyr)
library(stringr)
library(readr)
```

## Load it up

```{r}
fpkm = read_csv("../datasets/stratton_fpkm.csv")
glimpse(fpkm)
```

## Clean up

Let's get rid of X9 and X13, these are averages
```{r}
fpkm = fpkm %>% select(-X9, -X13)
```

and pull out the "Blood Aged" column, but we'll keep this in case we need it later
```{r}
blood = fpkm %>% select(`Blood Aged`)
fpkm = fpkm %>% select(-`Blood Aged`)
```


Now we can get rid of the first row 
```{r}
fpkm = fpkm %>% slice(-1)
```

## Getting variable information

Now we need to extract our variable information from the column names
```{r}
col_names = colnames(fpkm)[-1]

# get the day variable

day = str_extract(col_names, "day_\\d+")

# now the age variable

age = str_extract(col_names, "young|aged")

# we'll need a sample_ID variable for later
sample_ID = paste0("S", 1:length(age))
```

We'll create a new dataframe to hold our sample information, including a sample ID variable
```{r}
samples = data_frame(day, age, sample_ID)
```

now we rename our columns with the new sample ID, and renaming our gene column at the same time
```{r}
colnames(fpkm) = c("gene", sample_ID)
```


Now for the grand finale, joining the tables, first gathering the sample columns in the fpkm table and converting the expression column to numeric
```{r}
# first let's get rid of rows with NA's
fpkm = na.omit(fpkm)
fpkm_tidy = fpkm %>% 
	gather(sample_ID, expression, -gene) %>% 
	mutate(expression = as.numeric(expression))
final = left_join(samples, fpkm_tidy)
```

Tada!

## Making a heatmap

```{r}
library(pheatmap)
#set.seed(7592)
```

We'll need a matrix of expression values and a dataframe of sample info

```{r}

# filter out genes with all zeros
# get rid of duplicate genes which look to be an excel f*** up
# select 100 top genes

fpkm_small = fpkm_tidy %>% 
	group_by(gene) %>% 
	filter(sum(expression) != 0 ) %>% 
	group_by(sample_ID) %>% 
	filter(!duplicated(gene)) %>% 
	group_by(gene) %>% 
	mutate(mean_exp = mean(expression)) %>% 
	spread(sample_ID, expression) %>% ungroup() %>% 
	top_n(100, mean_exp)

exp = as.matrix(fpkm_small[,-c(1:2)])
rownames(exp) = fpkm_small$gene
```

Ok, let's make a heatmap

```{r}
pheatmap(exp)
pheatmap(log2(exp + 1))
```

Hmm, that's not right! Let's try scaling

```{r}
exp_log = log2(exp + 1)
pheatmap(exp_log, scale = "row")
```

Add an annotation to the columns
```{r}
col_anno = as.data.frame(samples[,-3])
rownames(col_anno) = samples$sample_ID

pheatmap(exp_log, annotation_col = col_anno, scale = "row")
```

`pheatmap` has loads of options.  Read the help page and play around

```{r}
our_gene = "Retnla"
head(final)

library(ggplot2)

final %>% filter(gene == our_gene) %>% 
	mutate(log_exp = log2(expression + 1)) %>% 
	ggplot(aes(x = day, y = log_exp)) +
	geom_boxplot(aes(fill = age)) +
	theme_bw() +
	scale_fill_brewer(palette = "Set1")
```


```{r}
sessionInfo()
```


