---
title: "PCA"
---

```{r, message=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(collapse = TRUE, message = FALSE, results = 'hide')
options("readr.num_columns" = 0)
```


```{r}
library("dplyr")
library("tidyr")
library("readr")
```


## PCA

### Getting the data ready
```{r}
# load the full dataset
pa_full = read_csv("../datasets/pa_phenotype_data.csv") %>% 
	mutate(Strain_ID = as.character(Strain_ID))

# select the columns we want and convert to a matrix in one step
pa_mat = pa_full %>% 
	select(C12_HSL, C4_HSL, Protease, Pyochelin, Pyoverdin, Swim, LB, M9glu, M9gly, SCFM) %>% 
	as.matrix()

# and we need some rownames
rownames(pa_mat) = pa_full$Strain_ID

# the basic PCA we're using can't handle missing values
pa_mat = na.omit(pa_mat)
```

### Do the PCA
```{r}
# run pca
pca = princomp(pa_mat)

# look at our results with basic built in plots
biplot(pca)

# hmmm, that doesn't look right, we must have forgotten to scale
# we want to scale rows so need to transpose and then transpose back for pca
pa_mat = t(scale(t(pa_mat)))

# try again
pca = princomp(pa_mat)
biplot(pca)
```

### A nicer plot

The base plot is not very nice, let's learn how to extract the info we want and make our own much nicer looking plot

#### Step 1: Extract the scores
```{r}
# pull them out of the pca object
scores = pca$scores

# we only want the first two principal components for plotting and let's rename them
scores = as.data.frame(scores[, 1:2])
colnames(scores) = c("PC1", "PC2")

# and finally we need to put the rownames back in because this is the StrainID
scores = scores %>% 
	tibble::rownames_to_column("Strain_ID")
```

#### Step 2: Combine back with original data

Here we'll use a join operation which will automatically find shared variable and join the two tables based on this variable.  In this case we'll use a left join so that any strains that are in the full set but not in the set we used for PCA are left out.

```{r}
pca_results = scores %>% left_join(pa_full)
```

#### Step 3: Make the plot

```{r}
pca_plot = pca_results %>% 
	ggplot(aes(x = PC1, y = PC2)) +
		geom_point(aes(colour = Clinical_Status)) +
	  theme_bw() +
	  scale_color_brewer(palette = "Set1")
pca_plot
```





