---
title: "Subsetting"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```


There are 6 different ways to subset and 3 subsetting operators.  Let's go through each data type and see how these work.  We'll start by using the `[` operator.

## Atomic vectors

```{r}
x = c(3.4, 5.3, 2.3, 1.0, 19.3)
```


Positive integers return elements at those positions

```{r}
x[c(2,4)]
```


Negative integers leave out elements at those positions

```{r}
x[-1]
x[-c(3,4)]
```

Don't mix positive and negative integers


Logical vectors return elements at the position where the value is `TRUE`. 

```{r}
x[c(TRUE, FALSE, TRUE, TRUE, FALSE)]
```

This is likely the method you'll use the most as you can write expressions that create a logical vector

```{r, collapse=TRUE}
x > 1
x[x > 1]
```


### Two less common methods

Nothing returns the original vector
```{r}
x[]
```

Zero returns a zero length vector

```{r}
x[0]
```


And finally if your vector is named you can use character vectors.  We'll use this a fair bit with data frames.
```{r}
y = setNames(x, c("A", "B", "C", "D", "E"))
y[c("C", "E", "A")]
```


## Lists

Lists work in the same way as vectors.  We'll look at how the different subsetting operators `[`, `[[`, and `$` change what you pull out of the list.

## Matrices

These concepts are applicable to higher-order arrays as well but we won't discuss those here.

```{r}
mat = matrix(1:9, nrow = 3)
colnames(mat) = c("A", "B", "C")
mat
```


We now have two dimensions, each one is subset like a one-dimensional vector. Rows are first, then columns.

```{r}
# keep all columns (the column index is blank)
mat[1:2, ]

# keep all rows (the row index is blank)
mat[, 2:3]

# both
mat[2:3, 1:2]

# use column names
mat[2:3, c("A", "B")]
```


Matrices can be indexed with a single vector and in this case the matrix will behave like a vector.

```{r}
mat

mat[1:4]
```

The matrix is stored in column-major order; count down the column first then start at the top of the next one.

## Data frames

Data frames can be indexed with a single vector like list (remember data frame columns are like a list).  They can (and are more commonly) indexed with two vectors like a matrix.

```{r}
df = data.frame(L = 1:3, M = 8:10, N = letters[2:4])

df
```

---- 

```{r}
df[c("M", "N")]

df[1:2, ]

df[1:2, c("L")]

df[df$M > 8, ]
```

---- 

A few more examples

```{r}
# predfined vector
rows_to_keep = sample(1:nrow(iris), size = 10)
new_iris = iris[rows_to_keep, ]
new_iris

# logical
new_iris$Species == "setosa"

new_iris[new_iris$Species == "setosa", ]
```

## Subsetting operators

* You can also `[[` and `$` to subset.  
* The difference is the type of object they return.  
* `[[` is most useful for lists because `[` always returns a list and not the actual value.  
* However, `[[` only returns a single value so it can only be used with a single positive integer or a string. `$` is short hand for `[[` with characters.


```{r, error=TRUE}
a = list(x = 1, y = 2)

typeof(a[1])

typeof(a[[1]])

a[[1:2]]

a[["y"]]
# or
a$y
```


The different operators are useful in different situations and you often find yourself using multiple operators in a single subsetting operation

```{r}
# $ is often used to pull out data frame columns or named list items
mtcars$cyl

# mix and match
mtcars[mtcars$cyl == 6, 1:4]

mtcars$mpg[mtcars$cyl == 6]

mtcars[mtcars$cyl == 6, c("mpg", "cyl", "hp")]
```


