---
title: "Subsetting"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```


There are 6 different ways to subset and 3 subsetting operators.  Let's go through each data type and see how these work.  We'll start by using the `[` operator.

## Atomic vectors

```{r}

```


Positive integers return elements at those positions

```{r}

```


Negative integers leave out elements at those positions

```{r}

```

Don't mix positive and negative integers


Logical vectors return elements at the position where the value is `TRUE`. 

```{r}

```

This is likely the method you'll use the most as you can write expressions that create a logical vector

```{r, collapse=TRUE}

```


### Two less common methods

Nothing returns the original vector
```{r}

```

Zero returns a zero length vector

```{r}

```


And finally if your vector is named you can use character vectors.  We'll use this a fair bit with data frames.
```{r}

```


## Lists

Lists work in the same way as vectors.  We'll look at how the different subsetting operators `[`, `[[`, and `$` change what you pull out of the list.

## Matrices

These concepts are applicable to higher-order arrays as well but we won't discuss those here.

```{r}

```


We now have two dimensions, each one is subset like a one-dimensional vector. Rows are first, then columns.

```{r}

```


Matrices can be indexed with a single vector and in this case the matrix will behave like a vector.

```{r}

```

The matrix is stored in column-major order; count down the column first then start at the top of the next one.

## Data frames

Data frames can be indexed with a single vector like list (remember data frame columns are like a list).  They can (and are more commonly) indexed with two vectors like a matrix.

```{r}

```

---- 

```{r}

```

---- 

A few more examples

```{r}

```

## Subsetting operators

* You can also `[[` and `$` to subset.  
* The difference is the type of object they return.  
* `[[` is most useful for lists because `[` always returns a list and not the actual value.  
* However, `[[` only returns a single value so it can only be used with a single positive integer or a string. `$` is short hand for `[[` with characters.


```{r, error=TRUE}

```


The different operators are useful in different situations and you often find yourself using multiple operators in a single subsetting operation

```{r}

```


